---
title: "Titanic SVM"
author: "Bhuvanesh Wadhwani"
date: "2023-05-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Calling packages 
```{r Packages, cache = TRUE}

packages_to_use<- c("ROCR", "tidyverse", "caret", "dplyr", "mice", "e1071", "ggplot2", "kernlab")

for(i in packages_to_use){
  if( ! i %in% rownames(installed.packages())  ) {
    print(paste(i, "not installed; installing now:\n") )
    install.packages(i)
  }
  
  require(i, character.only = TRUE)
}


```

# Part 1: Data Cleaning
#Taking in Datasets
```{r}

train.dat <- read.csv("train.csv")
test.dat <- read.csv("test.csv")

combined.dat <- bind_rows(train.dat, test.dat)

str(combined.dat)




```

pclass: A proxy for socio-economic status (SES)
1st = Upper
2nd = Middle
3rd = Lower
age: Age is fractional if less than 1. If the age is estimated, is it in the form of xx.5
sibsp: The dataset defines family relations in this way...
Sibling = brother, sister, stepbrother, stepsister
Spouse = husband, wife (mistresses and fiancés were ignored)
parch: The dataset defines family relations in this way...
Parent = mother, father
Child = daughter, son, stepdaughter, stepson
Some children travelled only with a nanny, therefore parch=0 for them.


```{r}
# Grab title from passenger names
combined.dat$Title <- gsub('(.*, )|(\\..*)', '', combined.dat$Name) #getting titles from name column into a new column 

# Show title counts by sex
table(combined.dat$Sex, combined.dat$Title)
```

```{r}
# Titles with very low cell counts to be combined to "rare" level
rare_title <- c('Dona', 'Lady', 'the Countess','Capt', 'Col', 'Don', 
                'Dr', 'Major', 'Rev', 'Sir', 'Jonkheer')

# Also reassign mlle, ms, and mme accordingly
combined.dat$Title[combined.dat$Title == 'Mlle']        <- 'Miss' 
combined.dat$Title[combined.dat$Title == 'Ms']          <- 'Miss'
combined.dat$Title[combined.dat$Title == 'Mme']         <- 'Mrs' 
combined.dat$Title[combined.dat$Title %in% rare_title]  <- 'Rare Title'

# Show title counts by sex again
table(combined.dat$Sex, combined.dat$Title)

```
```{r}
# Finally, grab surname from passenger name
combined.dat$Surname <- sapply(combined.dat$Name,  
                      function(x) strsplit(x, split = '[,.]')[[1]][1])
```


```{r}
# Create a family size variable including the passenger themselves
combined.dat$Fsize <- combined.dat$SibSp + combined.dat$Parch + 1

# Create a family variable 
combined.dat$Family <- paste(combined.dat$Surname, combined.dat$Fsize, sep='_')
```


```{r}
# Discretize family size
combined.dat$FsizeD[combined.dat$Fsize == 1] <- 'singleton'
combined.dat$FsizeD[combined.dat$Fsize < 5 & combined.dat$Fsize > 1] <- 'small'
combined.dat$FsizeD[combined.dat$Fsize > 4] <- 'large'

# Show family size by survival using a mosaic plot
mosaicplot(table(combined.dat$FsizeD, combined.dat$Survived), main='Family Size by Survival', shade=TRUE)
```


```{r}
# Create a Deck variable. Get passenger deck A - F:
combined.dat$Deck<-factor(sapply(combined.dat$Cabin, function(x) strsplit(x, NULL)[[1]][1]))
```

```{r}
# Get rid of our missing passenger IDs
embark_fare <- combined.dat %>%
  filter(PassengerId != 62 & PassengerId != 830)
```

```{r}
# Since their fare was $80 for 1st class, they most likely embarked from 'C'
combined.dat$Embarked[c(62, 830)] <- 'C'
```




```{r}
combined.dat$Fare[1044] <- median(combined.dat[combined.dat$Pclass == '3' & combined.dat$Embarked == 'S', ]$Fare, na.rm = TRUE)
```


```{r}
# Make variables factors into factors
factor_vars <- c('PassengerId','Pclass','Sex','Embarked',
                 'Title','Surname','Family','FsizeD')

combined.dat[factor_vars] <- lapply(combined.dat[factor_vars], function(x) as.factor(x))

# Set a random seed
set.seed(129)

# Perform mice imputation, excluding certain less-than-useful variables:
mice_mod <- mice(combined.dat[, !names(combined.dat) %in% c('PassengerId','Name','Ticket','Cabin','Family','Surname','Survived')], method='rf')
```
```{r}
# Save the complete output 
mice_output <- complete(mice_mod)
```


```{r}
# Replace Age variable from the mice model.
combined.dat$Age <- mice_output$Age

# Show new number of missing Age values
sum(is.na(combined.dat$Age))
```




```{r}
# Create the column child, and indicate whether child or adult
combined.dat$Child[combined.dat$Age < 18] <- 'Child'
combined.dat$Child[combined.dat$Age >= 18] <- 'Adult'

# Show counts
table(combined.dat$Child, combined.dat$Survived)
```



```{r}
# Adding Mother variable
combined.dat$Mother <- 'Not Mother'
combined.dat$Mother[combined.dat$Sex == 'female' & combined.dat$Parch > 0 & combined.dat$Age > 18 & combined.dat$Title != 'Miss'] <- 'Mother'

# Show counts
table(combined.dat$Mother, combined.dat$Survived)
```




```{r}
# Finish by factorizing our two new factor variables
combined.dat$Child  <- factor(combined.dat$Child)
combined.dat$Mother <- factor(combined.dat$Mother)
```


```{r}
md.pattern(combined.dat)

```


```{r}
# Split the data back into a train set and a test set
train.dat <- data.frame(combined.dat[1:891,])
test.dat <- data.frame(combined.dat[892:1309,])
```







#Factoring variables in the train dataset
```{r}
#Survived
train.dat$Survived <- factor(train.dat$Survived,
                       levels = c("0", "1"),
                       labels = c("No", "Yes"))

#Sex
train.dat$Sex <- factor(train.dat$Sex,
                       levels = c("female", "male"),
                       labels = c("Female", "Male"))

#Pclass
train.dat$Pclass <- factor(train.dat$Pclass,
                       levels = c("1", "2", "3"),
                       labels = c("Upper", "Middle", "Lower"))

#Embarked
train.dat$Embarked <- factor(train.dat$Embarked,
                       levels = c("C", "Q", "S"),
                       labels = c("Cherbourg", "Queenstown", "Southampton"))



```



#Factoring variables in the test dataset
```{r}
#Survived
test.dat$Survived <- factor(test.dat$Survived,
                       levels = c("0", "1"),
                       labels = c("No", "Yes"))

#Sex
test.dat$Sex <- factor(test.dat$Sex,
                       levels = c("female", "male"),
                       labels = c("Female", "Male"))

#Pclass
test.dat$Pclass <- factor(test.dat$Pclass,
                       levels = c("1", "2", "3"),
                       labels = c("Upper", "Middle", "Lower"))

#Embarked
test.dat$Embarked <- factor(test.dat$Embarked,
                       levels = c("C", "Q", "S"),
                       labels = c("Cherbourg", "Queenstown", "Southampton"))



```


# Part 2: SVM
```{r Train Control, cache = TRUE}
my.seed <- 111

set.seed(my.seed)
tr.Control <- trainControl(method = "repeatedcv",
                           number = 10,
                           repeats = 5,
                           classProbs = TRUE,
                           summaryFunction = twoClassSummary
                           )


```

# SVM Linear
```{r}
set.seed(111)
svm.linear <- train(Survived ~ Pclass + Sex + Age +SibSp + Parch + Fare + Embarked + Title + FsizeD + Child + Mother,
                    data = train.dat,
                    method = "svmLinear",
                    trControl = tr.Control,
                    metric = "ROC",
                    preProcess = c("center", "scale"),
                    tuneGrid = expand.grid(C = seq(.0001, 2, length = 30))
                    )

svm.linear$results 






```




```{r}

svm.linear$bestTune
#get the best measures:
best.index <- best(svm.linear$results, metric = "ROC", maximize = TRUE)
svm.linear$results[best.index, ]
#OR: 
svm.linear$results[ which.max(svm.linear$results$ROC), ]


svm.linear.pred <- predict(svm.linear, newdata = test.dat)

# Save the solution to a dataframe with two columns: PassengerId and Survived (prediction)
svm.sol <- data.frame(PassengerID = test.dat$PassengerId, Survived = svm.linear.pred)

# Write the solution to file
write.csv(svm.sol, file = 'svm_linear.csv', row.names = F)




```







# SVM Polynomial
```{r}

set.seed(111)
svm.poly <- train(Survived ~ Pclass + Sex + Age +SibSp + Parch + Fare + Embarked + Title + FsizeD + Child + Mother, 
              data = train.dat, 
              method = "svmPoly", 
              metric = "ROC",
              trControl = tr.Control, 
              preProcess = c("center","scale") 
              # , tuneLength = 15
              )
# the best tuning parameter
svm.poly$bestTune

svm.poly$results[ which.max(svm.poly$results$ROC), ]



svm.poly.pred <- predict(svm.poly, newdata = test.dat)

# Save the solution to a dataframe with two columns: PassengerId and Survived (prediction)
svm.pol <- data.frame(PassengerID = test.dat$PassengerId, Survived = svm.poly.pred)

# Write the solution to file
write.csv(svm.pol, file = 'svm_poly1.csv', row.names = F)


```



```{r}
plot(svm.poly)
```


```{r}
#grid search
set.seed(111) 
poly.grid <- expand.grid(C = .25,
                            scale = .001,
                         degree = seq(3, 5, length.out = 10))

set.seed(111)
svm.poly2 <- train(Survived ~ Pclass + Sex + Age + SibSp + Parch + 
                                            Fare + Embarked + Title + 
                                            FsizeD + Child + Mother, 
                      data = train.dat, 
                      method = 'svmPoly', 
                      trControl = tr.Control,
                      tuneGrid = poly.grid,
                      metric = "ROC",
                      preProcess = c("center", "scale")
                      )





# the best tuning parameter
svm.poly2$bestTune

svm.poly2$results[ which.max(svm.poly2$results$ROC), ]



svm.poly.pred2 <- predict(svm.poly2, newdata = test.dat)

# Save the solution to a dataframe with two columns: PassengerId and Survived (prediction)
svm.pol2 <- data.frame(PassengerID = test.dat$PassengerId, Survived = svm.poly.pred2)

# Write the solution to file
write.csv(svm.pol2, file = 'svm_poly2.csv', row.names = F)



```



# SVM Radial Basis Function Kernal
```{r}

set.seed(111)
svm.radial <- train(Survived ~ Pclass + Sex + Age +SibSp + Parch + Fare + Embarked + Title + FsizeD + Child + Mother, 
              data = train.dat, 
              method = "svmRadial", 
              metric = "ROC",
              trControl = tr.Control, 
              preProcess = c("center","scale")
              , tuneGrid = expand.grid(
                C = seq(0.01, 0.3, length.out = 10) 
                , sigma = c(0.01, 0.1, 1, 2)
                )
              )

svm.radial$bestTune
svm.radial$results
svm.radial$results[ which.max(svm.radial$results$ROC), ]




svm.rad.pred <- predict(svm.radial, newdata = test.dat)

# Save the solution to a dataframe with two columns: PassengerId and Survived (prediction)
svm.rad <- data.frame(PassengerID = test.dat$PassengerId, Survived = svm.rad.pred)

# Write the solution to file
write.csv(svm.rad, file = 'svm_rad1.csv', row.names = F)

```





```{r}

plot(svm.radial)


```



```{r}
set.seed(111)
svm.radial2 <- train(Survived ~ Pclass + Sex + Age +SibSp + Parch + Fare + Embarked + Title + FsizeD + Child + Mother, 
              data = train.dat, 
              method = "svmRadial", 
              metric = "ROC",
              trControl = tr.Control, 
              preProcess = c("center","scale")
              , tuneGrid = expand.grid(
                C = seq(0.01, 0.3, length.out = 10) 
                , sigma = seq(0.001, 0.1, length.out = 10)
                )
              )

svm.radial2$bestTune
svm.radial2$results
svm.radial2$results[ which.max(svm.radial2$results$ROC), ]




svm.rad.pred2 <- predict(svm.radial2, newdata = test.dat)

# Save the solution to a dataframe with two columns: PassengerId and Survived (prediction)
svm.rad2 <- data.frame(PassengerID = test.dat$PassengerId, Survived = svm.rad.pred2)

# Write the solution to file
write.csv(svm.rad2, file = 'svm_rad2.csv', row.names = F)
```





